Hydra Battles and Cie

### 2021-08-26 doc/part-hydras.tex
Alectryon:  first complete draft

\begin{Coqsrc}
Definition strongly_critical alpha := In (Cr alpha) alpha.

Definition maximal_critical alpha : Ensemble Ord :=
  fun gamma =>
    In (Cr alpha) gamma /\
    forall xi, alpha < xi -> ~ In (Cr xi) gamma.

Definition Gamma0 := the_least strongly_critical.
\end{Coqsrc}

@@

\noindent\emph{From \href{../theories/html/hydras.Gamma0.T2.html\#T2}%
{\texttt{Gamma0.T2}}}
\begin{Coqsrc}
(**  [gcons alpha beta n gamma] is : [psi(alpha,beta)*(S n)+ gamma]  *)

Inductive T2 : Set :=
| zero : T2
| gcons : T2 -> T2  -> nat -> T2 -> T2.

Notation "[ alpha , beta ]" := (gcons alpha beta 0 zero)
                                 (at level 0): t2_scope.
\end{Coqsrc}

@@

\begin{Coqsrc}
Notation  one  := [zero,zero].

(** The (n+1)-th finite ordinal *)
Notation FS n := (gcons zero zero n zero).

(** the [n]-th ordinal  *)
Definition fin (n:nat) := match n with 0 => zero | S p => FS p end.

Notation "'omega'"  := [zero,one] : T2_scope.
\end{Coqsrc}

\begin{Coqsrc}
Notation "'epsilon0'"  := ([one,zero]) : T2_scope.

Definition epsilon alpha := [one, alpha].
\end{Coqsrc}

@@

\begin{Coqsrc}
Inductive lt : T2 -> T2 -> Prop :=
| (* 1 *) 
 lt_1 : forall alpha beta n gamma,  zero t2< gcons alpha beta n gamma
| (* 2 *)
 lt_2 : forall alpha1 alpha2 beta1 beta2 n1 n2 gamma1 gamma2, 
                alpha1 t2< alpha2 ->
                beta1 t2< gcons alpha2 beta2 0 zero ->
               gcons alpha1 beta1 n1 gamma1 t2<
               gcons alpha2 beta2 n2 gamma2
| (* 3 *)
 lt_3 : forall alpha1  beta1 beta2 n1 n2 gamma1 gamma2, 
               beta1 t2< beta2 ->
               gcons alpha1 beta1 n1 gamma1 t2<
               gcons alpha1 beta2 n2 gamma2

| (* 4 *)
 lt_4 : forall alpha1 alpha2 beta1 beta2 n1 n2 gamma1 gamma2, 
               alpha2 t2< alpha1 ->
               [alpha1, beta1] t2< beta2 ->
               gcons alpha1 beta1 n1 gamma1 t2<
               gcons alpha2 beta2 n2 gamma2

| (* 5 *)
 lt_5 : forall alpha1 alpha2 beta1 n1 n2 gamma1 gamma2, 
               alpha2 t2< alpha1 ->
               gcons alpha1 beta1 n1 gamma1 t2<
               gcons alpha2  [alpha1, beta1] n2 gamma2

| (* 6 *)
 lt_6 : forall alpha1 beta1  n1  n2 gamma1 gamma2,  (n1 < n2)%nat ->
                                    gcons alpha1 beta1 n1 gamma1 t2< 
                                    gcons alpha1 beta1 n2 gamma2

| (* 7 *)
 lt_7 : forall alpha1 beta1 n1   gamma1 gamma2,  gamma1 t2< gamma2 ->
                                      gcons alpha1 beta1 n1 gamma1 t2<
                                      gcons alpha1 beta1 n1 gamma2
where  "o1 t2< o2" := (lt o1 o2): T2_scope.

Hint Constructors lt : T2.
\end{Coqsrc}

@@

\subsection{Examples}

\subsubsection*{Proof of $0<\epsilon_0$}

\begin{Coqsrc}
Example Ex1: 0 t2< epsilon0.
Proof.  constructor 1. Qed.
\end{Coqsrc}

\subsubsection*{Proof of $\omega<\epsilon_0$}

\begin{Coqsrc}
Example Ex2: omega t2< epsilon0.
Proof. info_auto with T2. (* uses lt_1 and lt_2 *) Qed.
\end{Coqsrc}

\subsubsection*{Proof of $\psi(\omega,8)\times 13+56 < \psi(\omega,8)\times 13+57 $}

\begin{Coqsrc}
Example Ex3: gcons omega 8 12 56 t2<  gcons omega 8 12 57.
Proof.
  constructor 7; constructor 6; auto with arith.
Qed.
\end{Coqsrc}

\subsubsection*{Proof of $\epsilon_0<\psi(2,1)$}

\begin{Coqsrc}
Example Ex4: epsilon0 t2< [2,1].
Proof.
   constructor 2; auto with T2.
   - constructor 6; auto with arith.
Qed.
\end{Coqsrc}

\subsubsection*{Proof of $\psi(2,1)<\psi(2,3)$}

\begin{Coqsrc}
Example Ex5 : [2,1] t2< [2,3].
Proof.
  constructor 3; auto with T2.
  - constructor 6; auto with arith.
Qed.
\end{Coqsrc}

\subsubsection*{Proof of $\psi(1,0)\times 13+ \omega < \psi(0,\psi(2,1))$}
\label{sect:ex6-first-proof}

\begin{Coqsrc}
Example Ex6 : gcons 1 0 12 omega t2< [0,[2,1]].
Proof.
  constructor 4.
  - constructor 1.
  - constructor 2.
    + constructor 6; auto with arith.
    + constructor 1.
Qed.
\end{Coqsrc}

\subsubsection*{Proof of $\psi(2,1)\times 43 + \epsilon_0 < \psi(1,\psi(2,1))$}

\begin{Coqsrc}
Example Ex7 : gcons 2 1 42 epsilon0 t2< [1, [2,1]].
Proof.
 constructor 5.
 constructor 6; auto with arith.
Qed.
\end{Coqsrc}

@@

\begin{Coqsrc}
Inductive nf : T2 -> Prop :=
| zero_nf : nf zero
| single_nf : forall a b n, nf a ->  nf b -> nf (gcons a b n zero)
| gcons_nf : forall a b n a' b' n' c', 
                      [a', b'] t2< [a, b]  -> 
                      nf a -> nf b -> 
                      nf(gcons a' b' n' c')-> 
                      nf(gcons a b n (gcons a' b' n' c')).
\end{Coqsrc}

@@

\begin{Coqsrc}
Lemma  nf_fin i : nf (fin i).
Proof.
  destruct i.
  - auto with T2.
  - constructor 2; auto with T2.
Qed.

Lemma nf_omega : nf omega.
Proof.  compute; auto with T2. Qed.

Lemma nf_epsilon0 : nf epsilon0.
Proof. constructor 2; auto with T2. Qed.

Lemma nf_epsilon : forall alpha, nf alpha -> nf (epsilon alpha).
Proof. compute; auto with T2. Qed.

Example Ex8: nf (gcons 2 1 42 epsilon0).
Proof.
  constructor 3; auto with T2.
  - apply Ex4.
  - apply nf_fin.
  - apply nf_fin.
Qed.
\end{Coqsrc}

@@

\begin{Coqsrc}
Fixpoint nbterms (t:T2) : nat :=
  match t with zero => 0
             | gcons a b n v => (S n) + nbterms v
  end.

Fixpoint t2_length (t:T2) : nat :=
  match t  with 
    zero => 0
  | gcons a b n v => 
       nbterms (gcons a b n v) + 
      2 * (Max.max (t2_length a)
                              (Max.max (t2_length b) 
                                                (t2_length_aux v)))
  end
with t2_length_aux (t:T2) : nat :=
 match t with 
 | zero => 0
  | gcons a b n v =>
           Max.max (t2_length a) 
                            (Max.max (t2_length b) (t2_length_aux v))
 end.
\end{Coqsrc}

\begin{Coqsrc}
Compute t2_length (gcons 2 1 42 epsilon0).
\end{Coqsrc}

\begin{Coqanswer}
 = 48 : nat
\end{Coqanswer}

@@

\begin{Coqsrc}
Lemma tricho_aux (l: nat) : forall t t' :T2,
      t2_length t + t2_length t' < l  ->
      {t t2< t'} + {t = t'} + {t' t2<  t}.
\end{Coqsrc}

Then we get our version of \texttt{lt\_eq\_lt\_dec}, and derive a comparison function;

\begin{Coqsrc}
Definition lt_eq_lt_dec (t t': T2) : {t t2< t'}+{t = t'}+{t' t2<  t}.
Proof.
  eapply tricho_aux.
  eapply lt_n_Sn.
Defined.

Definition compare (t1 t2 : T2) : comparison := 
  match lt_eq_lt_dec t1 t2 with
  | inleft (left _) => Lt
  | inleft (right _) => Eq
  | inright _ => Gt
  end.
\end{Coqsrc}

@@

\begin{Coqsrc}
Fixpoint nfb (alpha : T2) : bool :=
  match alpha with
    zero => true
  | gcons a b n zero => andb (nfb a) (nfb b)
  | gcons a b n ((gcons a' b' n' c') as c) =>
    match compare [a', b'] [a, b] with
           Lt => andb (nfb a) (andb (nfb b) (nfb c))
           | _ => false
           end
end.
\end{Coqsrc}

\begin{Coqsrc}
Compute compare (gcons 2 1 42 epsilon0) [2,2].
\end{Coqsrc}

\begin{Coqanswer}
   = Lt
     : comparison
\end{Coqanswer}

\begin{Coqsrc}
Compute nfb  (gcons 2 1 42 epsilon0).
\end{Coqsrc}

\begin{Coqanswer}
   = true
     : bool
\end{Coqanswer}

\begin{Coqsrc}
Compute nfb (gcons 2 1 42 (gcons 2 2 4 epsilon0)).
\end{Coqsrc}

\begin{Coqanswer}
   = false
     : bool
\end{Coqanswer}

\begin{remark}
The connection between the predicate \texttt{nf} and the relation \texttt{lt} on one part, and the functions \texttt{nfb} and \texttt{compare} on the other, is expressed by the following lemmas:

\begin{Coqsrc}
Lemma nfb_equiv gamma : nfb gamma = true <-> nf gamma.

Lemma compare_correct alpha beta :
  CompareSpec (alpha = beta) (lt alpha beta) (lt beta alpha)
              (compare alpha beta).
\end{Coqsrc}

The function \texttt{compare} helps to make easier proofs of inequalities of
closed terms of type \texttt{T2}.

First, we prove a lemma:

\begin{Coqsrc}
Lemma compare_Lt : forall alpha beta, compare alpha beta = Lt -> 
                                         alpha t2< beta.
Proof.
  intros alpha beta; destruct (compare_correct alpha beta);
    trivial; discriminate. 
Qed.
\end{Coqsrc}

Then, we give another version of the proof of Sect.~\vref{sect:ex6-first-proof}.

\begin{Coqsrc}
Example Ex6 : gcons 1 0 12 omega t2< [0,[2,1]].
Proof. now apply compare_Lt. Qed.
\end{Coqsrc}

@@

\noindent\emph{From \href{../theories/html/hydras.Gamma0.T2.html\#succ}%
{\texttt{Gamma0.T2}}}
\begin{Coqsrc}
Fixpoint succ (a:T2) : T2 :=
 match a with zero => one
             | gcons zero zero n c => fin (S (S n))
             | gcons a b n c => gcons a b n (succ c)
 end.
\end{Coqsrc}

@@

\begin{Coqsrc}
Fixpoint plus (t1 t2 : T2) {struct t1}:T2 :=
  match t1,t2 with
  |  zero, y  => y
  |  x, zero => x
  |  gcons a b n c, gcons a' b' n' c' =>
     (match compare (gcons a b 0 zero)
                    (gcons a' b' 0 zero) with
      | Lt => gcons a' b' n' c'
      | Gt => gcons a b n (c + gcons a' b' n' c')
      | Eq => gcons a b (S(n+n')) c'
      end)
  end
where "alpha + beta" := (plus alpha beta): T2_scope.
\end{Coqsrc}

\begin{Coqsrc}
Example Ex7 : 3 + epsilon0 = epsilon0.
Proof. trivial. Qed.
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition  phi (alpha beta : T2) : T2 :=
  match beta with zero => [alpha, beta] 
             | [b1, b2] => 
               (match compare alpha b1
                with Datatypes.Lt => [b1, b2 ]
                | _ => [alpha,[b1, b2]]
                end)
             | gcons b1 b2 0 (gcons zero zero  n zero) => 
               (match compare alpha b1
                with  Datatypes.Lt => 
                      [alpha, (gcons b1 b2 0 (fin n))]
                | _ =>  [alpha, (gcons b1 b2 0 (fin (S n)))]
                end)
             | any_beta => [alpha, any_beta]
  end.
\end{Coqsrc}

Despite its complexity, the function \texttt{phi} is well adapted to proofs by simplification or computation.
\begin{Coqsrc}
Example Ex8:  phi 1 (succ epsilon0) = [1, [1,0] + 1].
Proof. reflexivity. Qed.
\end{Coqsrc}

\begin{Coqsrc}
(**  All epsilons are fixpoints of phi 0 *)

Theorem epsilon_fxp : forall beta, phi zero (epsilon beta) =
                                   epsilon beta.
Proof. reflexivity. Qed.

Theorem epsilon0_fxp : epsilon0 = phi zero epsilon0.
Proof. apply epsilon_fxp. Qed.
\end{Coqsrc}

@@

\begin{Coqanswer}
phi_psi :
forall (alpha : T2) [beta gamma : T2] [n : nat],
nf gamma ->
limit_plus_fin beta n gamma ->
phi alpha beta = beta -> [alpha, gamma] = phi alpha (succ gamma)
\end{Coqanswer}

\begin{Coqsrc}
Example Ex9 : [zero, epsilon 2 + 4] = phi 0 (epsilon 2 + 5).
Proof. trivial. Qed.
\end{Coqsrc}

@@

\begin{Coqanswer}
phi_of_psi:
  forall a b1 b2 : T2,
  phi a [b1, b2] = (if lt_ge_dec a b1 then [b1, b2] else [a, [b1, b2]])
\end{Coqanswer}

\begin{Coqsrc}
Example Ex10 : phi omega [epsilon0, 5] = [epsilon0, 5].
Proof. reflexivity. Qed.
\end{Coqsrc}

@@

\begin{Coqsrc}
Module G0.

Class G0 := mkg0 {vnf : T2; vnf_ok : nfb vnf}.

Definition lt (alpha beta : G0) := T2.lt (@vnf alpha) (@vnf beta).

Definition compare alpha beta := Gamma0.compare (@vnf alpha) (@vnf beta).
\end{Coqsrc}

Then, we prove that \texttt{lt} is a well-founded strict order and that the
function \texttt{compare} is correct.

\begin{Coqsrc}
Instance lt_sto : StrictOrder lt.

Lemma lt_wf : well_founded lt.

Lemma compare_correct alpha beta :
  CompareSpec (alpha = beta) (lt alpha beta) (lt beta alpha)
              (compare alpha beta).

Instance Gamma0: ON lt  compare.
Proof.
  split.
  - apply lt_sto.
  - apply lt_wf. 
  - apply compare_correct.
Qed.
\end{Coqsrc}

@@

Prove that the implemantations of \texttt{succ}, \texttt{+}, $\phi_0$, etc.
are compatible in both notation systems.

@@

\noindent\emph{From \href{../theories/html/hydras.Gamma0.T2.html\#T1_to_T2}%
{\texttt{Gamma0.T2}}}
\begin{Coqsrc}
(* injection from T1 *)

Fixpoint T1_to_T2 (alpha :T1) : T2 :=
  match alpha  with
  | T1.zero => zero
  | T1.ocons a n b => gcons zero (T1_to_T2 a) n (T1_to_T2 b)
  end.
\end{Coqsrc}

>>>>

\input{movies/snippets/Critical/Gamma0Def}

@@

\noindent\emph{From \href{../theories/html/hydras.Gamma0.T2.html\#T2}%
{\texttt{Gamma0.T2}}}

@@

\input{movies/snippets/T2/Notations}

@@

\input{movies/snippets/T2/ltDef}

@@

\input{movies/snippets/T2/ltExamples}

@@

\input{movies/snippets/T2/nfDef}

@@

\input{movies/snippets/T2/nfExamples}

@@

@@

\input{movies/snippets/Gamma0/tricho}

\input{movies/snippets/Gamma0/compare}

@@

\input{movies/snippets/Gamma0/nfb}

@@

\noindent\emph{From \href{../theories/html/hydras.Gamma0.T2.html\#succ}%
{\texttt{Gamma0.T2}}}

\input{movies/snippets/T2/succDef}

@@

\noindent\emph{From \href{../theories/html/hydras.Gamma0.Gamma0T2.html\#succ}%
  {\texttt{Gamma0.Gamma0}}}

\input{movies/snippets/Gamma0/plusDef}

@@

\input{movies/snippets/Gamma0/phiDef}

Despite its complexity, the function \texttt{phi} is well adapted to proofs by simplification or computation.

@@

\input{movies/snippets/Gamma0/phiPsi}
\input{movies/snippets/Gamma0/Ex9}

@@

\input{movies/snippets/Gamma0/phiOfPsi}
\input{movies/snippets/Gamma0/Ex10}

@@

\input{movies/snippets/Gamma0/G0a}
\input{movies/snippets/Gamma0/G0b}

Then, we buils an instance of class \texttt{ON}. 
function \texttt{compare} is correct.

\input{movies/snippets/Gamma0/ltSto}
\input{movies/snippets/Gamma0/ltWf}
\input{movies/snippets/Gamma0/ONGamma0}

@@

Prove that the implementations of \texttt{succ}, \texttt{+}, $\phi_0$, etc.
are compatible in both notation systems.

@@

\noindent\emph{From \href{../theories/html/hydras.Gamma0.T2.html\#T1_to_T2}%
{\texttt{Gamma0.T2}}}

\input{movies/snippets/T2/T1ToT2}
